The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-11-03 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Spell check

\section{Introduction}
In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based or satellite transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequence of this approach is that a certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed. It is important to analyze the impact of RFI on a specific instrument to optimize the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Obervatory in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.

@@

\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citeauthor{lofar-radio-environment} (\citeyear{lofar-radio-environment}; hereafter referred to as \citetalias{lofar-radio-environment}) it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and conclude in \citet{offringa-rfi-distributions} that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

\begin{figure}
\begin{center}\hspace*{-0.8cm}\includegraphics[width=10.5cm]{img/bandpass}
\caption{Correlator output RMS over frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data. In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimize quantization noise. The 1.28~MHz pattern is due to the imperfect poly-phase filter.
}
\label{fig:bandpass}
\end{center}
\end{figure}

To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying fields of view. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citetalias{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ on one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging for getting accurate flux calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

@@

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not significantly cause any time or bandwidth decorrelation up to the first beam nul. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $uvw$ values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation.

@@

\section{Data description} \label{ch:data-description}
We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The nights of observations are listed in Table~\ref{tbl:obs-list}.

@@

Both projects avoid the subbands that cover the frequency range 133.8--138.9~MHz, because the ORBCOMM low Earth orbitting (LEO) satellites transmit in these subbands. When these frequencies are observed, the subbands are often so strongly contaminated that it affects imaging sensitivity. For completeness, we include a 16-min observation that covers this frequency range.

The poly-phase filter bank that performs the first separation into subbands, introduces a 1.28~MHz periodic spectral signature. This shape is e.g. visible in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

@@

Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital TV (DTV), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.

@@

Although the flagger does not adequately flag DTV, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DTV interference from the data statistics. An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in DTV band 4, which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%. The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz. Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth. The visibility ratios that are lost because of DTV RFI are detailed in the last column of Table~\ref{tbl:obs-list}. Because these frequencies correspond to the EoR redshift, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.

@@

\noindent\begin{figure*}
\subcaptionbox{RFI contamination found in the 2 m amateur band (146 MHz).}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\label{fig:amateur-2m}
}\\\vspace{3mm}%
\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. Detection of this kind of RFI requires flagging at high time and frequency resolution.}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/150_2_mhz_example_ann}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged_ann}
\label{fig:150_2}
}%
\caption{RFI events found in the EoR low band (138.9--169.6 MHz) in a 2-min snapshot with relatively high RFI contamination. These panels show the Stokes I amplitudes. In the right figures, the result of RFI detection is shown with purple. The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
\label{fig:2m-amateur-band}
\end{figure*}

@@

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DTV RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines is added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyze these phenomena.

@@

\section{Comparison with LOFAR} \label{ch:lofar-comparison}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citetalias{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency subrange of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citetalias{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.

@@

Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI levels. Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction. These values can therefore be interpreted and compared as being the minimum loss of data due to RFI.

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar computational requirements as for LOFAR. Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site. However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

@@

SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.

The \textsc{aoflagger} RFI detection strategy, originally developed for LOFAR, works well for the MWA. Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events. These algorithms have gained some popularity (e.g., besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in Miriad), but other projects still use single-sample thresholding, e.g. PAPER \citep{parsons-paper-eorlimit-2014}. Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory. Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.

>>>>

\section{Introduction}
In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequence of this approach is that a certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed. It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Observatory in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.

@@

\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citeauthor{lofar-radio-environment} (\citeyear{lofar-radio-environment}; hereafter referred to as \citetalias{lofar-radio-environment}) it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

\begin{figure}
\begin{center}\hspace*{-0.8cm}\includegraphics[width=10.5cm]{img/bandpass}
\caption{Correlator output RMS over frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data. In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimise quantisation noise. The 1.28~MHz pattern is due to the imperfect poly-phase filter.
}
\label{fig:bandpass}
\end{center}
\end{figure}

To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows customising its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying fields of view. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citetalias{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ on one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging for getting accurate flux calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimising reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

@@

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not significantly cause any time or bandwidth decorrelation up to the first beam null. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $uvw$ values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation.

@@

\section{Data description} \label{ch:data-description}
We analyse the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The nights of observations are listed in Table~\ref{tbl:obs-list}.

@@

Both projects avoid the sub-bands that cover the frequency range 133.8--138.9~MHz, because the ORBCOMM low Earth orbiting (LEO) satellites transmit in these sub-bands. When these frequencies are observed, the sub-bands are often so strongly contaminated that it affects imaging sensitivity. For completeness, we include a 16-min observation that covers this frequency range.

The poly-phase filter bank that performs the first separation into sub-bands, introduces a 1.28~MHz periodic spectral signature, which is shown in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

@@

Fig.~\ref{fig:rfi-per-band} shows the overall detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital TV (DTV), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.

@@

Although the flagger does not adequately flag DTV interference, the right plot of Fig.~\ref{fig:dvb} shows that from the data statistics it is possible to determine whether a snapshot is affected by DTV interference. An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in DTV band 4, which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%. The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz. Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth. The visibility ratios that are lost because of DTV RFI are detailed in the last column of Table~\ref{tbl:obs-list}. Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.

@@

\noindent\begin{figure*}
\subcaptionbox{RFI contamination found in the 2 m amateur band (146 MHz).}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\label{fig:amateur-2m}
}\\\vspace{3mm}%
\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. Detection of this kind of RFI requires flagging at high time and frequency resolution.}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/150_2_mhz_example_ann}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged_ann}
\label{fig:150_2}
}%
\caption{RFI events found in the EoR low band (138.9--169.6 MHz) in a 2-min snapshot with relatively high RFI contamination. These panels show the Stokes I amplitudes. In the right figures, the result of RFI detection is shown with purple. The horizontal flagged lines are flagged because they are W-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
\label{fig:2m-amateur-band}
\end{figure*}

@@

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occasionally, DTV RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines is added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyse these phenomena.

@@

\section{Comparison with LOFAR} \label{ch:lofar-comparison}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citetalias{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citetalias{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.

@@

Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI levels. Nevertheless, for both telescopes, the automatic detection strategies have been optimised such that as little data as possible are thrown away, but to be sufficient for further data reduction. These values can therefore be interpreted and compared as being the minimum loss of data due to RFI.

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR. Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site. However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

@@

SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.

The \textsc{aoflagger} RFI detection strategy, originally developed for LOFAR, works well for the MWA. Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events. These algorithms have gained some popularity (e.g., besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}), but other projects still use single-sample thresholding, e.g. PAPER \citep{parsons-paper-eorlimit-2014}. Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory. Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.
